---
title: 'Introduction to cartography'
author: "Timothée Giraud, Nicolas Lambert"
date: '`r Sys.Date()`'
output: 
  rmarkdown::html_vignette:
    toc: true
    number_sections: true 
    toc_depth: 1
vignette: >
  %\VignetteIndexEntry{Introduction to cartography}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}

---



```{r echo=FALSE}
knitr::opts_chunk$set(collapse = TRUE)

knitr::knit_hooks$set(margin = function(before, options, envir){
  if (before){
    par(mar=c(0.1,0.1,1.3,0.1))
  } 
})
```

# Introduction

The aim of `cartography` is to obtain thematic maps with the visual quality of those build with other common mapping and GIS software.  

Users of the package could belong to one of two categories: cartographers willing to use R or R users willing to create maps. Therefore, its functions have to be intuitive to cartographers and ensure compatibility with common R workflows.  

`cartography` uses `sf` or `sp` objects to produces `base` graphics. As most of the internals of the package relies on `sf` functionalities, the preferred format for spatial objects is `sf`.  


# Features
`cartography`'s functions can be classified in the following categories :

* **Symbology** (*`*Layer()` functions*)     
Each function focuses on a single cartographic representation (e.g. proportional symbols or choropleth representation) and displays it on a georeferenced plot. This solution allows to consider each representation as a layer and to overlay multiple representations on a same map.  
Each function has two main arguments that are:

    - `x`, a spatial object (preferably an `sf` object),  
    - `var`, the name of a variable to map.  

    `sp` objects are handled through the `spdf` argument if the variable is contained within the `Spatial*DataFrame` and through `spdf`, `spdfid`, `df`, `dfid` if the variable is in a separate `data.frame` that needs to be joined to the `Spatial*DataFrame`. 

    Many parameters are available to fine tune the cartographic representation. These parameters are the common ones found in GIS and automatic cartography software (e.g. classification and color palettes used in choropleth maps, symbols sizes used in proportional symbols maps...).

* **Legends** (*`legend*()` functions*)     
Legends are displayed by default along cartographic layers but more parameters are available through `legend*()` functions.

* **Transformations** (*`get*()` functions*)    
A set of functions is dedicated to the creation or transformation of spatial objects (e.g. borders extraction, grid or links creation). These functions are provided to ease the creation of some more advanced maps that usually need geo-processing. 

* **Map Layout**    
Along with the cartographic functions, some other functions are dedicated to layout design (e.g. customizable scale bar, north arrow, title, sources or author information...).

* **Color Palettes**     
16 original color palettes are shipped within the package. Those palettes can be customized and combined. 



# Examples


## OpenStreetMap Base Map and Proportional Symbols

`getTiles()` and `tilesLayer()` download and display OpenStreetMap tiles. Be careful to cite the source of the tiles appropriately. 

`propSymbolsLayer()` displays symbols proportionals to a quantitative variable (stocks). Several symbols are available (circles, squares, bars). The `inches` argument is used to customize the symbols sizes. 

```{r propMap, fig.height=6, fig.width=5, message=FALSE, margin=TRUE}
library(sf)
library(cartography)
# path to the geopackage file embedded in cartography
path_to_file <- system.file("gpkg/mtq.gpkg", package="cartography")
# import to an sf object
mtq <- st_read(dsn = path_to_file, quiet = TRUE)
# download osm tiles
mtq.osm <- getTiles(
  x = mtq, 
  type = "osm", 
  zoom = 11, 
  crop = TRUE
)
# plot osm tiles
tilesLayer(x = mtq.osm)
# plot municipalities (only borders are plotted)
plot(st_geometry(mtq), col = NA, border = "grey", add=TRUE)
# plot population
propSymbolsLayer(x = mtq, var = "POP", inches = 0.4, col = "brown4",
  legend.pos = "topright",  legend.title.txt = "Total population")
# layout
layoutLayer(title = "Population Distribution in Martinique",
            sources = "Sources: Insee and IGN - 2018\n© OpenStreetMap contributors.\nTiles style under CC BY-SA, www.openstreetmap.org/copyright.",
            author = paste0("cartography ", packageVersion("cartography")),
            frame = FALSE, north = FALSE, tabtitle = TRUE)
# north arrow
north(pos = "topleft")
```


## Choropleth Map

In choropleth maps, areas are shaded according to the variation of a quantitative variable. They are used to represent ratios or indices. 

`choroLayer()` displays choropleth maps . Arguments `nclass`, `method` and `breaks` allow to customize the variable classification. `getBreaks()` allow to classify outside of the function itself. Colors palettes are defined with `col` and a set of color palettes can be created with `carto.pal()` (see also `display.carto.all()`). 


```{r choroMap, fig.height=6, fig.width=5, margin=TRUE}
library(sf)
library(cartography)
# path to the geopackage file embedded in cartography
path_to_file <- system.file("gpkg/mtq.gpkg", package="cartography")
# import to an sf object
mtq <- st_read(dsn = path_to_file, quiet = TRUE)
# population density (inhab./km2) using sf::st_area()
mtq$POPDENS <- 1e6 * mtq$POP / st_area(mtq)
# plot municipalities (only the backgroung color is plotted)
plot(st_geometry(mtq), col = NA, border = NA, bg = "lightblue1")
# plot population density
choroLayer(
  x = mtq, 
  var = "POPDENS",
  method = "geom",
  nclass=5,
  col = carto.pal(pal1 = "sand.pal", n1 = 5),
  border = "grey40", 
  lwd = 0.5,
  legend.pos = "topright", 
  legend.title.txt = "Population Density\n(people per km2)",
  add = TRUE
) 
# http://webhelp.esri.com/arcgisdesktop/9.2/index.cfm?topicname=geometrical_interval
# layout
layoutLayer(title = "Population Distribution in Martinique", 
            sources = "Sources: Insee and IGN - 2018",
            author = paste0("cartography ", packageVersion("cartography")), 
            frame = FALSE, north = FALSE, tabtitle = TRUE) 
# north arrow
north(pos = "topleft")

```


## Colored Pencil Maps and Typolgies


```{r pentypoMap, fig.height=6, fig.width=5, margin=TRUE}
library(sf)
library(cartography)
# path to the geopackage file embedded in cartography
path_to_file <- system.file("gpkg/mtq.gpkg", package="cartography")
# import to an sf object
mtq <- st_read(dsn = path_to_file, quiet = TRUE)
# transform municipality multipolygons to (multi)linestrings
mtq_pencil <- getPencilLayer(
  x = mtq, 
  size = 500, 
  lefthanded = F
)
# plot municipalities (only the backgroung color is plotted)
plot(st_geometry(mtq), col = NA, border = NA, bg = "lightblue1")
# plot administrative status
typoLayer(
  x = mtq_pencil, 
  var="STATUS",  
  col = c("aquamarine4", "yellow3","wheat"), 
  lwd = .7,
  legend.values.order = c("Prefecture",
                          "Sub-prefecture", 
                          "Simple municipality"),
  legend.pos = "topright",
  legend.title.txt = "", 
  add=TRUE
)
#  plot municipalities
plot(st_geometry(mtq), lwd = 0.5, border = "grey20", add = TRUE, lty = 3)
# labels for a few  municipalities
labelLayer(x = mtq[mtq$STATUS != "Simple municipality",], txt = "LIBGEO", 
           cex = 0.9, halo = T, r = 0.15)
# north arrow
north(pos = "topleft", col = "#EED19A")
# title, source, author
layoutLayer(title = "Administrative Status", 
            sources = "Sources: Insee and IGN - 2018",theme = "sand.pal",
            author = paste0("cartography ", packageVersion("cartography")), 
            frame = FALSE, north = FALSE, tabtitle = TRUE) 
```

## Label Map

`labelLayer()` is dedicated to the display of labels on a map. The `overlap = FALSE` argument displays non overlapping labels. 

```{r labelMap, fig.height=6, fig.width=5, margin=TRUE}
library(sf)
library(cartography)
# path to the geopackage file embedded in cartography
path_to_file <- system.file("gpkg/mtq.gpkg", package="cartography")
# import to an sf object
mtq <- st_read(dsn = path_to_file, quiet = TRUE)
# plot municipalities
plot(st_geometry(mtq), col = "darkseagreen3", border = "darkseagreen4", 
     bg = "lightblue1", lwd = 0.5)
# plot labels
labelLayer(
  x = mtq, 
  txt = "LIBGEO", 
  col= "black", 
  cex = 0.7, 
  font = 4,
  halo = TRUE, 
  bg = "white", 
  r = 0.1, 
  overlap = FALSE, 
  show.lines = FALSE
)
# map layout
layoutLayer(
  title = "Municipalities of Martinique", 
  sources = "Sources: Insee and IGN - 2018",  
  author = paste0("cartography ", packageVersion("cartography")), 
  frame = FALSE,
  north = TRUE, 
  tabtitle = TRUE
) 
```



## Isopleth Map

Unlike choropleth maps, which represent data as discrete values, isopleth maps are based on the assumption that the phenomenon to be represented has a continuous distribution. These maps use a spatial interaction modeling approach which aims to compute indicators based on stock values weighted by distance (Stewart 1942). It allows a spatial representation of the phenomenon independent from the initial heterogeneity of the territorial division. The result is easy to read and can be considered as a bypassing of the Modifiable Areal Unit Problem (MAUP) (Openshaw and Taylor 1979).

In cartography, the smoothLayer function, that heavily depends on the SpatialPosition (Giraud and Commenges 2016) package, allow to compute quickly these maps. The function takes as inputs a marked point layer and a set of parameters (a spatial interaction function and its parameters) and outputs an isopleth map layer.

```{r isopleth, fig.height=6, fig.width=5, margin=TRUE}
library(sf)
library(cartography)
# path to the geopackage file embedded in cartography
path_to_file <- system.file("gpkg/mtq.gpkg", package="cartography")
# import to an sf object
mtq <- st_read(dsn = path_to_file, quiet = TRUE)
# plot municipalities (only the backgroung color is plotted)
plot(st_geometry(mtq), col = NA, border = NA, bg = "lightblue1")
# plot isopleth map
smoothLayer(
  x = mtq, 
  var = 'POP',
  span = 4000,
  nclass = 12,
  beta = 2,
  mask = mtq, 
  border = "grey",
  lwd = 0.1, legend.values.rnd = -3,
  col = carto.pal(pal1 = 'brown.pal', n1 = 12),
  legend.title.txt = "Population\nPotential",
  legend.pos = "topright", add=TRUE
)
# annotation on the map
text(x = 692582, y = 1611478, cex = 0.8, adj = 0, font = 3,  labels = 
       "Distance function:\n- type = exponential\n- beta = 2\n- span = 4 km")
# layout
layoutLayer(title = "Population Distribution in Martinique",
            sources = "Sources: Insee and IGN - 2018",
            author = paste0("cartography ", packageVersion("cartography")),
            frame = FALSE, north = FALSE, tabtitle = TRUE)
# north arrow
north(pos = "topleft")

```


## Grid Map

The grid-cell method is an option to overcome the arbitrariness and irregularity of administrative division. It highlights the main trends in the data spatial distribution, splitting the territory in regular blocks. Statistical values are distributed over a regular grid. Cell values are classified and then displayed in areas of color. The principle adopted here is to set each cell’s value with a proportion of the initial geometrical units it overlay (share of intersected area).

In cartography the map process relies on three functions. getGridLayer builds a regular grid (squares or hexagons) based on a spatial object and provide a table of surface intersections. getGridLayer computes data to match the grid layer according to the surface intersections. Eventually, choroLayer uses the grid and data to display a choropleth map (figure 4).


```{r grid, fig.height=6, fig.width=5, margin=TRUE}
library(sf)
library(cartography)
# path to the geopackage file embedded in cartography
path_to_file <- system.file("gpkg/mtq.gpkg", package="cartography")
# import to an sf object
mtq <- st_read(dsn = path_to_file, quiet = TRUE)

# Create a grid layer
mygrid <- getGridLayer(
  x = mtq, 
  cellsize = (median(st_area(mtq))), 
  var = c("POP"),
  type = "hexagonal"
)
# Compute population density in people per km2
mygrid$POPDENS <- 1e6 * mygrid$POP / mygrid$gridarea
# plot municipalities (only the backgroung color is plotted)
plot(st_geometry(mtq), col = NA, border = NA, bg = "lightblue1")
# Plot the population density
choroLayer(x = mygrid, var = "POPDENS", method = "geom", nclass=5, 
           col = carto.pal(pal1 = "turquoise.pal", n1 = 5), border = "grey80", 
           lwd = 0.5, legend.pos = "topright", add = TRUE,
           legend.title.txt = "Population Density\n(people per km2)") 
layoutLayer(title = "Population Distribution in Martinique", 
            sources = "Sources: Insee and IGN - 2018",
            author = paste0("cartography ", packageVersion("cartography")), 
            frame = FALSE, north = FALSE, tabtitle = TRUE) 
```



## Discontinuities

Discontinuities maps are based on the variation of a phenomena between contiguous administrative units. This kind of representation does not focus on homogeneous zones, but rather on spatial breaks. On the map, discontinuity intensity is expressed by the borders’ thickness.

The first step to build these maps is to extracts borders between units. The second step is to compute a discontinuity measure (either a ratio or an absolute difference). The third step is to display it on a map using the size visual variable. Combine these discontinuities with a choropleth representation helps to understand the discontinuity direction (which one of two regions has the higher value).

In cartography, getBorders is used to build a spatial object of borders between units. Each resulting borders contains the ids of its two neighboring units. It is also possible to complement these borders by the getOuterBorders function that can compute borders between non-contiguous units (maritime borders). discLayer function compute and display discontinuities, the line widths reflect the ratio or the absolute difference between values of an indicator in two neighboring units. Of course, the discontinuities layer can be associate with a choropleth layer (choroLayer) (figure 5).


```{r discc, fig.height=5, fig.width=5, margin=TRUE}
library(sf)
library(cartography)
# path to the geopackage file embedded in cartography
path_to_file <- system.file("gpkg/mtq.gpkg", package="cartography")
# import to an sf object
mtq <- st_read(dsn = path_to_file, quiet = TRUE)
# Compute the population density (inhab./km2) using sf::st_area()
mtq$POPDENS <- as.numeric(1e6 * mtq$POP / st_area(mtq))
# Get a SpatialLinesDataFrame of countries borders
mtq.contig <- getBorders(mtq)

plot(st_geometry(mtq), col = NA, border = NA, bg = "lightblue1", 
     xlim = c(690574, 745940))
# Plot the population density
choroLayer(x = mtq, var = "MED",
           breaks = c(min(mtq$MED), seq(13000, 21000, 2000), max(mtq$MED)),
           col = carto.pal("green.pal", 6),border = "white", lwd = 0.5, 
           legend.pos = "topright", legend.title.txt = "Median Income\n(euros)",
           add = TRUE)

# Plot discontinuities
discLayer(
  x = mtq.contig, 
  df = mtq, 
  var = "MED",
  type = "rel", 
  method = "geom", 
  nclass = 3,
  threshold = 0.4,
  sizemin = 0.7, 
  sizemax = 6, 
  col = "red4",
  legend.values.rnd = 1, 
  legend.title.txt = "Relative\nDiscontinuities", 
  legend.pos = "right",
  add = TRUE
)

# Layout
layoutLayer(title = "Wealth Disparities in Martinique, 2015", 
            author =  paste0("cartography ", packageVersion("cartography")),
            sources = "Sources: Insee and IGN - 2018",
            frame = FALSE, scale = 5, tabtitle = TRUE,theme = "grey.pal")

```


## `sp` objects


```{r sp, fig.height=5, fig.width=7, margin=TRUE}
library(sp)
library(cartography)
data("nuts2006")


# Plot a layer with the extent of the EU28 countries with only a background color
plot(nuts0.spdf, border = NA, col = NA, bg = "#A6CAE0")
# Plot non european space
plot(world.spdf, col  = "#E3DEBF", border = NA, add = TRUE)
# Plot Nuts2 regions
plot(nuts0.spdf, col = "grey60",border = "white", lwd = 0.4, add = TRUE)

propSymbolsLayer(
  spdf = nuts0.spdf, 
  df = nuts0.df, 
  spdfid = "id", 
  dfid = "id", 
  var = "pop2008", 
  legend.pos = "topright", 
  col = "red4", 
  border = "white", 
  legend.title.txt = "Population" 
)

layoutLayer(title = "Population in Europe, 2008",
            sources = "Data: Eurostat, 2008",
            author =  paste0("cartography ", packageVersion("cartography")),
            scale = 500, frame = TRUE, col = "#688994") 

north("topleft")
```





# Datasets

Several datasets are embedded in the package: 

- A [GeoPackage](https://www.geopackage.org/) of Martinique municipalities can be imported via the `st_read()` function of the `sf` package.   
  - Sources: [Base comparateur de territoires](https://www.insee.fr/fr/statistiques/2521169) (upload date: 2018-09-25) & [ADMIN EXPRESS-COG](http://professionnels.ign.fr/adminexpress) (2018 edition). 
  - Citation: Insee and IGN, 2018
  - Fields: 
    - **INSEE_COM**: Municipality identifier
    - **STATUS**: Municipality administrative status
    - **LIBGEO**: Municipality name
    - **POP**: Total population, 2015
    - **MED**: Median disposable income adjusted per equivalent household member, in euros, 2015
    - **CHOM**: Unemployed population, 2015
    - **ACT**: Active population, 2015
- A csv file of professional mobility flows between Martinique municipalities. 
  - Sources: [Flux de mobilité - déplacements domicile-lieu de travail](https://www.insee.fr/fr/statistiques/3566477) (upload date: 2018-08-01) 
  - Citation: ©Insee, 2018
  - Fields: 
    - **i**: Municipality of residence identifier
    - **j**: Municipality of workplace identifier
    - **fij**: Flows of workers (employed population, 15 y.o. or more, 2015, only flows > 100) 
- A set of `sp` objects and `data.frame`s on European regions (NUTS) can be loaded in the environment via `data(nuts2006)`. Each layer of this dataset is directly described in the documentation (e.g. `?nuts0.spdf`).  


